### 2015-12-11 paper/2-background.tex

\subsection{Optimistic Synchronization}
A completely different synchronization protocol is \textit{optimistic synchronization}~\cite{TimeWarp}.
Whereas conservative synchronization would prevent causality errors at all costs, optimistic synchronization will allow them to happen, but correct them afterwards.
Each node is allowed to progress in simulated time as much as possible, without taking note of the state of any other node.
When an event arrives at a node, which is already further in simulated time, the node will have to roll back its state to right before the event would normally have to be processed.
As the simulation time is now rolled back to before the event is processed, the event can simply be processed as if no causality error ever occured.

Rolling back the simulation time requires the node to store past model states, such that they can be restored later on.
Furthermore, all incoming and outgoing events need to be stored as well.
Incoming events need to be passed to the models again, when the correct simulation time has again been reached, and outgoing events need to be cancelled, as potentially a different series of output events would normally have been generated.
Cancelling events, however, can cause further rollbacks, as the receiving node might also have to roll back its state.
In practice, a single causality error could have significant repercussions on the complete simulation.

Further changes are required for unrecoverable operations, such as I/O (\textit{e.g.}, tracing, writing to file, printing output) and memory management.
Lightweight algorithms are still required to determine the lower bound of all simulation times, through the computation of a \textit{Global Virtual Time} (GVT).

The main advantage is that performance is not limited by a small lookahead, caused by a very infrequent event.
If an (almost) instantaneous event occurs rarely, performance will only be impacted if it occurs, and not at every simulation step.
The main disadvantage is unpredictable performance and arbitrary cost of rollbacks due to the propagation of causality errors.
If rollbacks occur frequently, simulation quickly becomes slow, as the overhead of the recovery mechanisms becomes significant.
Apart from overhead in CPU time, a significant memory overhead is present: all intermediate states qre stored up to a point where it can be considered \textit{irreversible}.

While optimistic synchronization does not explicitly depends on the lookahead, simulation performance is still bound by the lookahead implicitly.

>>>>

\subsection{Optimistic Synchronization}
A completely different synchronization protocol is \textit{optimistic synchronization}~\cite{TimeWarp}.
Whereas conservative synchronization prevents causality errors at all costs, optimistic synchronization allows them to happen, but corrects them.
Each node simulates as fast as possible, without taking note of any other node.
It assumes that no events occur from other nodes, unless it has explicitly received one at that time.
When this assumption is violated, the node rolls back its simulation time and state to right before the moment when the event has to be processed.
As simulation time is rolled back to before where the event must be processed, the event can just be processed as if no causality error ever occured.

Rolling back simulation time requires the node to store past model states, such that they can be restored later.
All incoming and outgoing events need to be stored as well.
Incoming events are injected again after a rollback, when their time has been reached again.
Outgoing events are cancelled after a rollback, through the use of anti-messages, as potentially different output events have to be generated.
Cancelling events, however, can cause further rollbacks, as the receiving node might also have to roll back its state.
In practice, a single causality error can have significant repercussions.

Further changes are required for unrecoverable operations, such as I/O and memory management.
These are only executed after the lower bound of all simulation times, called \textit{Global Virtual Time} (GVT), has progressed beyond their execution time.

The main advantage is that performance is not limited by a small lookahead, caused by a very infrequent event.
If an (almost) instantaneous event rarely occurs, performance is only impacted when it occurs, and not at every simulation step.
The disadvantage is unpredictable performance due to the arbitrary cost of rollbacks and their propagation.
If rollbacks occur frequently, state saving and rollback overhead can cause simulation to grind to a halt.
Apart from overhead in CPU time, a significant memory overhead is present: all intermediate states are stored up to a point where it can be considered \textit{irreversible}.

Note that, while optimistic synchronization does not explicitly depends on lookahead, performance still implicitly depends on lookahead.
